Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-04 paper/zhang-cook-laa-langrene-menendez.tex
changes on some highlighted places and a minor tweak of search plot

Several R implementations address optimization problems with both
general purpose as well as task specifics solvers. The most prominent
one within the general solvers is \code{optim()} in the \CRANpkg{stats}
\citep{stats} package, which provides both gradient-based and
derivative-free optimisation functions. Another general solver
specialised in non-linear optimisation is \CRANpkg{nloptr}
\citep{nloptr}. Specific solvers for simulated annealing includes
\code{optim(..., method  "SANN")} and package \CRANpkg{GenSA}
\citep{gensa} that deals with more complicated objective functions. For
other task specific solvers, readers are recommended to visit the
relevant sections in CRAN task review on
\ctv{optimisation and mathematical programming}
\citep{crantaskreviewoptim}.

@@

Projection pursuit guided tour combines two different methods
(projection pursuit and guided tour) in exploratory data analysis.
Projection pursuit, coined by \citet{friedman1974projection}, detects
interesting structures (e.g.~clustering, outliers and skewness) in
multivariate data via low dimensions projection. Guided tour is one
variation of a broader class of data visualisation methods, tour, which
displays high-dimensional data through a series of animated projections.

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A d-dimensional projection is a linear
transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\) defined as
\(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto an scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. A number of index functions have been proposed in
the literature to detect different data structures
\citep[friedman1974projection;][]{hall1989polynomial, cook1993projection, posse1995projection}
and to classify groups \citep{lee2005projection, lee2010projection}.

As a general visualisation method, tour produces animations of high
dimensional data via rotations of low dimension planes. There are
different tour implementations depending on how the high dimensional
space is investigated: grand tour \citep{cook2008grand} selects the
planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.

@@

{\centering \includegraphics[width=0.5\linewidth,height=0.2\textheight]{/Users/pmen0008/Documents/paper-tour-vis/img/tour-path}

@@

The optimisation problem in projection pursuit is stated as follows:
Given a randomly start basis \(\mathbf{A}_1\), projection pursuit finds
the final projection basis \(\mathbf{A}_T\) that satisfies the following
optimisation problem:

@@

\noindent where \(I_d\) is the \(d\)-dimensional identity matrix and the
constraint requires the projection bases \(\mathbf{A}\) to be
orthonormal. Several features of this optimisation are worth-noticing.
1) The optimisation is constrained and the space of the bases is a
d-dimensional sphere. 2) The objective function may not be
differentiable for a constructed index function. 3) Although finding the
global maximum is the goal of an optimisation problem, it is also
interesting to inspect local maximum in projection pursuit since it
could present unexpected interesting projections. 4) The optimisation
procedure needs to be easy to compute since the tour animation needs to
be played in real-time during the optimisation.

@@

Three main pieces of information are recorded for the projection pursuit
optimisers: 1) projection bases \(\mathbf{A}\), 2) index values \(I\),
and 3) state \(S\). For SA and SAJO, possible state include
\texttt{random\_search}, \texttt{new\_basis}, and
\texttt{interpolation}. Pseudo derivative (PD) has a wider variety of
state including \texttt{new\_basis}, \texttt{direction\_search},
\texttt{best\_direction\_search}, \texttt{best\_line\_search}, and
\texttt{interpolation}. Multiple iterators index the information
collected at different levels: \(t\) is a unique identifier prescribing
the natural ordering of each observation; \(j\) and \(l\) are the
counter of the outer and inner loop respectively. Other parameters of
interest recorded include \texttt{method} that tags the name of the
optimiser, and \texttt{alpha} that indicates the sampling neighbourhood
size for searching observations. A matrix notation of the data structure
is presented in Equation \ref{eq:data-structure}.

@@

\noindent where \(T^{\prime} = T + k_{J}+ l_{J+1}\). Note that there is
no output in iteration \(J + 1\) since the optimiser doesn't find a
better basis in the last iteration and terminates. The final basis found
is \(A_T\) with index value \(I_T\).

@@

A starting point of diagnosing an optimiser is to understand how many
search an optimiser has conducted. One may want to simply plot the index
value of the search points across its natural order, but a point
geometry may work well if each iteration has a similar number of points.
When some iterations have considerably more points than others, using a
point geometry over-emphasizes the iterations that have more search
points since these iterations will occupy the mast majority of the plot
space. An alternative is to summarise the search in each iteration using
boxplot and each iteration will then be spaced out equally.
Occasionally, one may still want to switch back to a point geometry if
the number of points is small in a particular iteration and this can be
achieved via the \texttt{cutoff} argument in the search plot function.
Additional annotations are added to facilitate better reading of the
plot and these includes\\
1) the number of points searched in each iteration can be added as text
label at the bottom of each iteration; 2) the anchor bases to
interpolate are connected and highlighted in a larger size; and 3) the
colour of the last iteration is in a grey scale to indicate no better
basis found in this iteration.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-search-1}

@@

\caption[A comparison of search by two otimisers]{A comparison of search by two otimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iteration because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-interp-1}

@@

\caption[Trace plots of the interpolating bases with the same optimisation routine as the previous figure]{Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotinic.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

Another interest of diagnostic is to examine how the index value
improves for interpolating bases since the projection on these bases is
played by the tour animation. Trace plots are created with plotting the
index value of the interpolation points against time. Figure
\ref{fig:toy-interp} presents the trace plot of the same optimisation
routine as Figure \ref{fig:toy-search}. The trace plot is smooth and an
interesting discovery is that in SAJO, the interpolation in iteration
6-8 is not monotonic. Sometimes, this non-monotonicity is acceptable
since the target basis does not necessarily have a higher index value
than the current one, while in some cases, the optimisation will be
better off if these intermediate basis can be used as the start of the
next iteration (see section \ref{monotonic} for a detailed discussion on
this).

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-pca-1}

@@

Apart from checking the search and progression of an optimiser, looking
at where the bases are positioned in the basis space is another
interest. Given the orthonormality constraint, the space of projection
bases \(\mathbf{A}_{p \times d}\) is a \(p \times d\) dimension sphere
and dimension reduction methods, i.e.~principal component analysis is
applied to first project all the bases onto a 2D space. In a projection
pursuit guided tour optimisation, there are various type of bases
involved: 1) The start basis; 2) The search bases that an optimiser
evaluated to produce the anchor bases; 3) The anchor bases that have the
highest index value in each iteration; 4) The interpolating bases on the
interpolation path; and finally 5) the end basis. The importance of
these basis differs and the most important ones are the start,
interpolating and end bases. Anchor and search bases can be turned on
with argument \texttt{details\ =\ TRUE}. Sometimes, two optimisers can
start with the same basis but finish with bases of opposite sign. This
happens because the projection is invariant to the sign difference of
the bases and so does the index value, however, this creates
difficulties for comparing the optimisers since the end bases will be
symmetric with respect to the origin. A sign flipping step is conducted
to flip the signs of all the bases in one routine if different
optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As
mentioned previously, the original basis space is a high dimensional
sphere and random bases on the sphere can be generated via the CRAN
package \CRANpkg{geozoo} \citep{geozoo}. Along with the bases recorded
during the optimisation and a zero basis, the parameters (centre and
radius) of the 2D space can be estimated. PCA is performed to get the
first two PC coordinates of all the bases and the centre of the circle
is the PC coordinates of the zero matrix, and radius is estimated as the
largest distance from the centre to the basis. The theoretical best
basis is known for simulated data and can be labelled to compare the how
close the end basis found by the optimisers is to the theoretical best.
Various aesthetics, i.e.~size, alpha and colour, are applicable to
emphasize critical elements and adjust for the presentation. For
example, anchor points and search points are less important and hence a
smaller size and alpha is used; Alpha can also be applied on the
interpolation paths to show the start to finish from transparent to
opaque.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-pca-animated-1}

@@

Animation is another display to show how the search progresses from
start to finish in the space and an \texttt{animate\ =\ TRUE} argument
is used to enable the animation in the PCA plot. Figure
\ref{fig:toy-pca-animated} shows six frames from the animation of the
PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information
one can learn from this animation is that SA finds its end basis quicker
than PD since SA finishes its search in the 5th frame while PD is still
making more progression.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/toy-tour-1}

@@

\caption{Six selected frames from the tour animation for viewing the same two optmisations as previous two figures. The tour animation allows for an appreciation of the search bases in the original high dimensional space. A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

As mentioned previously, the original \(p \times d\) dimension space can
be simulated via randomly generated bases in \CRANpkg{geozoo}
\citep{geozoo} package. While the PCA plot projects the bases from the
direction that maximises the variance, the tour plot display of the
original high dimensional space from various directions using animation.
Figure \ref{fig:toy-tour} shows frames from the tour plot of the same
two optimisations in its original 5D space.

@@

Random variables with different distributions have been simulated and
the distributional form of each variable is presented in Equations
\ref{eq:sim-norm} to \ref{eq:sim-x7}. Variable \texttt{x1}, \texttt{x8},
\texttt{x9} and \texttt{x10} are normal noise with zero mean and unit
variance and \texttt{x2} to \texttt{x7} are normal mixtures with varied
weights and locations. All the variables have been scaled to have an
overall unit variance before projection pursuit.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/polish-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/flip-sign-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/noisy-better-geo-1}

@@

\caption[Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimiers]{Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimiers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO success in optimise the index, SAJO takes much longer than SA and finishes off closer to the theoretial best.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of
all three optimisers and as expected, none of the interpolated path is
smooth. There is barely any improvement made by PD, indicating its
failure in optimising non-smooth index. While SA and SAJO have managed
to get close to the index value of the theoretical best, trace plot
shows that it takes SAJO much longer to interpolate towards the final
basis. This long interpolation path is partially due to the fluctuation
in the early iterations, where the SAJO tends to generously accept
inferior bases before concentrating near the optimum. The PCA plot shows
the interpolation path and search points excluding the last termination
iteration. {[}A sentence for PD{]}. Comparing the amount of random
search done by SA and SAJO, it is surprising that SAJO doesn't carry
that many of sampling as SA. Combine the insight from both the trace and
PCA plot, one can learn the two different search strategies by SA and
SAJO: SA frequently samples in the space and only make a move when an
improvement is guaranteed to be made, while SAJO first broadly accepts
bases in the space and then starts the extensive sampling in a narrowed
subspace. The specific decision of which optimiser to use will depend on
the index curve in the basis space but if the basis space is non-smooth,
accepting inferior bases at first, as what SAJO has done, can lead to a
more efficient search, in terms of the overall number of points
evaluated.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Documents/paper-tour-vis/figs/kol-result-1}

@@

\caption[PCA plot of 1D projection on the 6\-variable data, \code{boa6}, with normal kolmogorov index]{PCA plot of 1D projection on the 6\-variable data, \code{boa6}, with normal kolmogorov index. Two optimisers: SA and SAJO and two search sizes: 0.5 and 0.7 are used in the simulation for 20 replicates. Local optimum (x), is presented in this experiment, along with the global optimum (*). The final basis is larger and more opaque than the start basis for each path and color represents whether the global or local optimum is found, after polishing.}\label{fig:kol-result}
\end{figure}
\end{Schunk}

The next experiment compares the performance of SA and SAJO when a local
maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared
to understand how a large search neighbourhood would affect the final
basis and the length of the search. Figure \ref{fig:kol-result} shows 80
paths simulated using 20 seeds in the PCA plot, faceted by optimiser and
search size. With SA and a search size of 0.5, despite being the
simplest and fastest, the optimiser fails in three instances where the
final basis lands neither near the local nor the global optimum. With a
larger search size of 0.7, more seeds have found the global maximum.
Comparing SA and SAJO for a search size of 0.5, SAJO doesn't seem to
improve the final basis found, despite having longer interpolation
paths. However, the denser paths near the local maximum is an indicator
that SAJO is working hard to examine if there is any other optimum in
the basis space but the relative small search size has diminished its
ability to reach the global maximum. With a larger search size, almost
all the seeds (16 out of 20) have found the global maximum and some
final bases are much closer to the theoretical best, as compared to the
three other cases. This indicates that SAJO, with a reasonable large
search size, is able to overcome the local optimum and optimise close
towards the global optimum.

@@

  \begin{itemize}
  \tightlist
  \item
    \code{theme\_fern()} and \code{format\_label()} for better display
    of the grid lines and axis formatting
  \item
    \code{clean\_method()} for clean up the name of the optimisers
  \item
    \code{botanical\_palettes} is a collection of colour palettes from
    Australian native plants Quantitative palettes include daisy,
    banksia and cherry and sequential palettes contain fern and acacia
  \item
    \code{botanical\_pal()} as the colour interpolator
  \item
    \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the
    colour and fill of the plot
  \end{itemize}
\end{itemize}

@@

Most of the work in this project has been translated into code in two
packages: the collection of the data object is implemented in the
existing \CRANpkg{tourr}\citep{tourr} package; manipulation and
visualisation of the data object are implemented in the new \pkg{ferrn}
package. Equipped with handy tools to diagnose the performance of
optimisers, future work can extend the diagnostics to a wider range of
index functions, i.e.~scagnostics, association, and information index
\citep{laa2020using} and understand how the optimisers behave for index
functions with different structures.

@@

This article is created using \CRANpkg{knitr}\citep{knitr} and
\CRANpkg{rmarkdown} \citep{rmarkdown} in R. The source code for
reproducing this paper can be found at:
\url{https://github.com/huizezhang-sherry/paper-tour-vis}.

>>>>

Several R implementations address optimization problems with both
general purpose as well as task specifics solvers. The most prominent
one within the general solvers is \code{optim()} in the \CRANpkg{stats}
\citep{stats} package, which provides both gradient-based and
derivative-free optimisation functions. Another general solver
specialised in non-linear optimisation is \CRANpkg{nloptr}
\citep{nloptr}. Specific solvers for simulated annealing include
\code{optim(..., method  "SANN")} and package \CRANpkg{GenSA}
\citep{gensa} that deals with more complicated objective functions. For
other task specific solvers, readers are recommended to visit the
relevant sections in CRAN task review on
\ctv{optimisation and mathematical programming}
\citep{crantaskreviewoptim}.

@@

Projection pursuit guided tour combines two different methods
(projection pursuit and guided tour) in exploratory data analysis.
Projection pursuit, coined by \citet{friedman1974projection}, detects
interesting structures (e.g.~clustering, outliers and skewness) in
multivariate data via low-dimensional projections. Guided tour is one
variation of a broader class of data visualisation methods, tour, which
displays high-dimensional data through a series of animated projections.

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. A number of index functions have been proposed in
the literature to detect different data structures
\citep{friedman1974projection, hall1989polynomial, cook1993projection, posse1995projection}
and to classify groups \citep{lee2005projection, lee2010projection}.

As a general visualisation method, tour produces animations of high
dimensional data via rotations of low-dimensional planes. There are
different tour implementations depending on how the high dimensional
space is investigated: grand tour \citep{cook2008grand} selects the
planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.

@@

{\centering \includegraphics[width=0.5\linewidth,height=0.2\textheight]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/img/tour-path}

@@

In projection pursuit the optimisation problem aims at finding the
global optimum that gives the most interesting projection according to
an index function. That is, it starts with a given randomly selected
basis \(\mathbf{A}_1\) and aims at finding an optimal final projection
basis \(\mathbf{A}_T\) in each iteration that satisfies the following
optimisation problem:

@@

\noindent where \(f\), \(\mathcal{A}\), and \(\mathbf{X}\) are defined
as in the previous section, and \(I_d\) is the \(d\)-dimensional
identity matrix and the constraint requires the projection bases
\(\mathbf{A}\) to be orthonormal. Several features of this optimisation
are worth-noticing. 1) The optimisation is constrained and the space of
the bases is a \(d\)-dimensional sphere. 2) The objective function may
not be differentiable for a constructed index function. 3) Although
finding the global maximum is the goal of an optimisation problem, it is
also interesting to inspect local maximum in projection pursuit since it
could present unexpected interesting projections. 4) The optimisation
procedure needs to be easy to compute since the tour animation needs to
be played in real-time during the optimisation.

@@

Three main pieces of information are recorded for the projection pursuit
optimisers: 1) projection bases \(\mathbf{A}\), 2) index values \(I\),
and 3) state \(S\). For SA and SAJO, possible states include
\texttt{random\_search}, \texttt{new\_basis}, and
\texttt{interpolation}. Pseudo derivative (PD) has a wider variety of
states including \texttt{new\_basis}, \texttt{direction\_search},
\texttt{best\_direction\_search}, \texttt{best\_line\_search}, and
\texttt{interpolation}. Multiple iterators index the information
collected at different levels: \(t\) is a unique identifier prescribing
the natural ordering of each observation; \(j\) and \(l\) are the
counter of the outer and inner loop respectively. Other parameters of
interest recorded include \texttt{method} that tags the name of the
optimiser, and \texttt{alpha} that indicates the sampling neighbourhood
size for searching observations. A matrix notation of the data structure
is presented in equation \ref{eq:data-structure}.

@@

\noindent where \(T^{\prime} = T + k_{J}+ l_{J+1}\). Note that there is
no output in iteration \(J + 1\) since the optimiser does not find a
better basis in the last iteration and terminates. The final basis found
is \(A_T\) with index value \(I_T\).

@@

A starting point of diagnosing an optimiser is to understand how many
searches an optimiser has conducted. One may want to simply plot the
index value of the search points across its natural order, but a point
geometry may work well if each iteration has a similar number of points.
When some iterations have considerably more points than others, using a
point geometry over-emphasizes the iterations that have more search
points since these iterations will occupy the mast majority of the plot
space. An alternative is to summarise the search in each iteration using
boxplots and each iteration will then be spaced out equally.
Occasionally, one may still want to switch back to a point geometry if
the number of points is small in a particular iteration and this can be
achieved via the \texttt{cutoff} argument in the search plot function.
Additional annotations are added to facilitate better reading of the
plot and these include\\
1) the number of points searched in each iteration can be added as text
label at the bottom of each iteration; 2) the anchor bases to
interpolate are connected and highlighted in a larger size; and 3) the
colour of the last iteration is in a grey scale to indicate no better
basis found in this iteration.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1}

@@

\caption[A comparison of search by two optimisers]{A comparison of search by two optimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iterations because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1}

@@

\caption[Trace plots of the interpolating bases with the same optimisation routine as the previous figure]{Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotonic.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

Another interesting diagnostic is to examine how the index value
improves between interpolating bases since the projection on these bases
is played by the tour animation. Trace plots are created with plotting
the index value of the interpolation points against time. Figure
\ref{fig:toy-interp} presents the trace plot of the same optimisation
routine as Figure \ref{fig:toy-search}. The trace plot is smooth and an
interesting discovery is that in SAJO, the interpolation in iteration
6-8 is not monotonic. Sometimes, this non-monotonicity is acceptable
since the target basis does not necessarily have a higher index value
than the current one, while in some cases, the optimisation will be
better off if these intermediate basis can be used as the start of the
next iteration (see section \ref{monotonic} for a detailed discussion on
this).

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-1}

@@

Apart from checking the search and progression of an optimiser, looking
at where the bases are positioned in the basis space is another
interest. Given the orthonormality constraint, the space of projection
bases \(\mathbf{A}_{p \times d}\) is a \(p \times d\)-dimensional
sphere. A dimensionality reduction method, e.g.~principal component
analysis is applied to first project all the bases onto a 2D space. In a
projection pursuit guided tour optimisation, there are various types of
bases involved: 1) The starting basis; 2) The search bases that the
optimiser evaluated to produce the anchor bases; 3) The anchor bases
that have the highest index value in each iteration; 4) The
interpolating bases on the interpolation path; and finally 5) the end
basis. The importance of these bases differs and the most important ones
are the starting, interpolating and end bases. The anchor and search
bases can be turned on with argument \texttt{details\ =\ TRUE}.
Sometimes, two optimisers can start with the same basis but finish with
bases of opposite sign. This happens because the projection is invariant
to the sign difference of the bases and so does the index value,
however, this creates difficulties for comparing the optimisers since
the end bases will be symmetric with respect to the origin. A sign
flipping step is conducted to flip the signs of all the bases in one
routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As
mentioned previously, the original basis space is a high-dimensional
sphere and random bases on the sphere can be generated via the CRAN
package \CRANpkg{geozoo} \citep{geozoo}. Along with the bases recorded
during the optimisation and a zero basis, the parameters (centre and
radius) of the 2D space can be estimated. PCA is performed to get the
first two PC coordinates of all the bases and the centre of the circle
is the PC coordinates of the zero matrix, and radius is estimated as the
largest distance from the centre to the basis. The theoretical best
basis is known for simulated data and can be labelled to compare how
close the end basis found by the optimisers is to the theoretical best.
Various aesthetics, i.e.~size, alpha and colour, are applicable to
emphasize critical elements and adjust for the presentation. For
example, anchor points and search points are less important and hence a
smaller size and alpha is used. Alpha can also be applied on the
interpolation paths to show the start to finish from transparent to
opaque.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1}

@@

Animation is another type of display to show how the search progresses
from start to finish in the space. An \texttt{animate\ =\ TRUE} argument
is used to enable the animation in the PCA plot. Figure
\ref{fig:toy-pca-animated} shows six frames from the animation of the
PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information
one can learn from this animation is that SA finds its end basis quicker
than PD since SA finishes its search in the 5th frame while PD is still
making more progress.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1}

@@

\caption{Six selected frames from the tour animation for viewing the same two optimisations as previous two figures. The tour animation allows for an appreciation of the search bases in the original high dimensional space. A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

As mentioned previously, the original \(p \times d\) dimension space can
be simulated via randomly generated bases in the \CRANpkg{geozoo}
\citep{geozoo} package. While the PCA plot projects the bases from the
direction that maximises the variance, the tour plot displays the
original high-dimensional space from various directions using animation.
Figure \ref{fig:toy-tour} shows some frames from the tour plot of the
same two optimisations in its original 5D space.

@@

Random variables with different distributions have been simulated and
the distributional form of each variable is presented in equations
\ref{eq:sim-norm} to \ref{eq:sim-x7}. Variables \texttt{x1},
\texttt{x8}, \texttt{x9} and \texttt{x10} are normal noise with zero
mean and unit variance and \texttt{x2} to \texttt{x7} are normal
mixtures with varied weights and locations. All the variables have been
scaled to have an overall unit variance before projection pursuit.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/flip-sign-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1}

@@

\caption[Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimisers]{Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimisers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO succeed in optimising the index, SAJO takes much longer than SA and finishes off closer to the theoretial best.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of
all three optimisers and as expected, none of the interpolated path is
smooth. There is barely any improvement made by PD, indicating its
failure in optimising non-smooth index. While SA and SAJO have managed
to get close to the index value of the theoretical best, trace plot
shows that it takes SAJO much longer to interpolate towards the final
basis. This long interpolation path is partially due to the fluctuation
in the early iterations, where the SAJO tends to generously accept
inferior bases before concentrating near the optimum. The PCA plot shows
the interpolation path and search points excluding the last termination
iteration. \textcolor{red}{[A sentence for PD]}. Comparing the amount of
random search done by SA and SAJO, it is surprising that SAJO does not
carry as many samples as SA. Combining the insights from both the trace
and PCA plot, one can learn the two different search strategies by SA
and SAJO: SA frequently samples in the space and only make a move when
an improvement is guaranteed to be made, while SAJO first broadly
accepts bases in the space and then starts the extensive sampling in a
narrowed subspace. The specific decision of which optimiser to use will
depend on the index curve in the basis space but if the basis space is
non-smooth, accepting inferior bases at first, as what SAJO has done,
can lead to a more efficient search, in terms of the overall number of
points evaluated.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-result-1}

@@

\caption[PCA plot of 1D projection on the 6\-variable data, \code{boa6}, with normal kolmogorov index]{PCA plot of 1D projection on the 6\-variable data, \code{boa6}, with normal kolmogorov index. Two optimisers: SA and SAJO and two search sizes: 0.5 and 0.7 are used in the simulation for 20 replicates. The local optimum (x), is presented in this experiment, along with the global optimum (*). The final basis is larger and more opaque than the start basis for each path and color represents whether the global or local optimum is found, after polishing.}\label{fig:kol-result}
\end{figure}
\end{Schunk}

The next experiment compares the performance of SA and SAJO when a local
maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared
to understand how a large search neighbourhood would affect the final
basis and the length of the search. Figure \ref{fig:kol-result} shows 80
paths simulated using 20 seeds in the PCA plot, faceted by optimiser and
search size. With SA and a search size of 0.5, despite being the
simplest and fastest, the optimiser fails in three instances where the
final basis lands neither near the local nor the global optimum. With a
larger search size of 0.7, more seeds have found the global maximum.
Comparing SA and SAJO for a search size of 0.5, SAJO does not seem to
improve the final basis found, despite having longer interpolation
paths. However, the denser paths near the local maximum is an indicator
that SAJO is working hard to examine if there is any other optimum in
the basis space but the relative small search size has diminished its
ability to reach the global maximum. With a larger search size, almost
all the seeds (16 out of 20) have found the global maximum and some
final bases are much closer to the theoretical best, as compared to the
three other cases. This indicates that SAJO, with a reasonable large
search size, is able to overcome the local optimum and optimise close
towards the global optimum.

@@

  \begin{itemize}
  \tightlist
  \item
    \code{theme\_fern()} and \code{format\_label()} for better display
    of the grid lines and axis formatting
  \item
    \code{clean\_method()} to clean up the name of the optimisers
  \item
    \code{botanical\_palettes} is a collection of colour palettes from
    Australian native plants. Quantitative palettes include daisy,
    banksia and cherry and sequential palettes contain fern and acacia
  \item
    \code{botanical\_pal()} as the colour interpolator
  \item
    \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the
    colour and fill of the plot
  \end{itemize}
\end{itemize}

@@

Most of the work in this project has been translated into code in two
packages: the collection of the data object is implemented in the
existing \CRANpkg{tourr} \citep{tourr} package; while the manipulation
and visualisation of the data object are implemented in the new
\pkg{ferrn} package. Equipped with handy tools to diagnose the
performance of optimisers, future work can extend the diagnostics to a
wider range of index functions, i.e.~scagnostics, association, and
information index \citep{laa2020using} and understand how the optimisers
behave for index functions with different structures.

@@

This article is created using \CRANpkg{knitr} \citep{knitr} and
\CRANpkg{rmarkdown} \citep{rmarkdown} in R. The source code for
reproducing this paper can be found at:
\url{https://github.com/huizezhang-sherry/paper-tour-vis}.
